Grade Retention and Mixed Race

### 2022-11-11 products/paper/main.qmd
Clean out most TODO notes from paper draft

The experiences of mixed race individuals will play an important role in the future of race in the US. A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines [@root_racially_1992]. This demographic phenomenon may have a potentially transformative effect on racial boundaries [@hochschild_creating_2012; @alba_great_2020]. While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for its demographic size and for a post-Civil Rights environment in which historical norms of racial classification may be eroding. Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into a prevailing racial hierarchy that reinforces a system of racial inequality.

Extensive scholarship documents the identification and classification decisions made by or on behalf of multiracial individuals [@xie_racial_1997; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009] <!-- TODO MORE HERE -->. This research has generally found a surprising degree of fluidity and choice in classification decisions <!-- TODO CHECK -->. On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers [@kao_racial_1999; @campbell_multiracial_2009; @herman_blackwhiteother_2009]. Two important data limitations have contributed to the lack of work on this important topic.

@@

One important outcome, grade retention, is measurable and important even for young children, however. Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race <!-- TODO CITATIONS -->. Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience and within it we can clearly see the imprint of the racial hierarchy. In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers. Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents.

@@

How might we expect the outcomes for biracial individuals do differ from their monoracial peers? In every case, the relevant comparison is how a biracial person fares relative to the two constituent groups with which their parents identify. For example, in the case of an individual with one Black and one White parent, we would compare the average outcomes for such individuals to the average outcomes for individuals with two White and two Black parents, respectively. The outcome for the Black/White group may be similar to one of the monoracial groups, it may be somewhere in between the outcomes for the two monoracial groups, or it may be higher or lower than either of the monoracial groups.

@@

Discrimination and bias by race play an important role in overall levels of racial inequality. Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowningly rely upon and reinforce in social interactions <!-- TODO citations, may be Pager, Brubaker? -->. Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers. Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into organizations and institutional systems <!-- TODO citations, Tilly? Bonilla-Silva, Pager again? -->. In our current "colorblind" regime, such systems may be formally race neutral but produce disparate impacts by race <!-- TODO CITE Bonilla-Silva? -->.

@@

The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm concerning Black ancestry. According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries[@davis_who_1991]. @davis_who_1991 generalizes this practice into the concept of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group. By comparison a pattern of *hyperdescent* implies that individuals of mixed racial ancestry will be classified as members of the higher status group. This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry <!-- TODO CITE -->.

Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals <!-- TODO citations -->, there are reasons to be cautious about the generalizability of this paradigm to both the contemporary and historical situation. Hypodescent and hyperdescent are observed in two specific historical cases involving Black ancestry and American Indian ancestry, respectively. As @iverson_regimes_2022 note, these two cases could just as parsimoniously be explained by a "dominance" model which ranks ancestries by their tendencies to be supercessive or recessive in determining identification and classification. In this framework, within the historical "racial triangle" of American colonialism, Black ancestry is supercessive, Indigenous ancestry is recessive, and White ancestry falls somewhere in between.

The development of these particular norms may also reflect the historical regime in which they were developed [@gullickson_choosing_2011; @iverson_regimes_2022]. In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us to fully understand the experiences of those of Latino and Asian ancestry due to more recent processes of racialization. Second, the historical norms for all groups may also be breaking up or transforming in the post-Civil Rights era <!-- TODO CITE -->.

@@

Racial inequality is also at least partly explained by different distributions of material, cultural, and social resources across racial groups <!-- TODO CITATIONS HERE? -->. On average, socioeconomic background (e.g. parental income, wealth, and education) differs by race and such socioeconomic background is consequential for outcomes. Other less materially defined resources also vary across groups such as cultural styles, social networks, and neighborhood quality. When such characteristics are held constant or "controlled", the remaining differences between racial groups tend to be smaller, although such differences rarely disappear <!-- TODO WHAT CAN I CITE HERE -->. These differences in background resources reflect the accumulation of historical processes of discrimination <!-- CITE PAGER? -->. Thus, observed racial inequality reflects both contemporary processes and the intergenerational accumulation of disadvantage.

@@

Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage [@davis_intermarriage_1941; @merton_intermarriage_1941; @fu_racial_2001]. Using education as the socioeconomic parameter, prior research has shown strong evidence of this pattern in the case of Black male/White female marriages in the US, but weaker evidence for other types of unions [@gullickson_education_2006; @fu_racial_2001; @kalmijn_educational_2010; @hou_interracial_2011]. Less research has explored these patterns outside of marriage <!-- TODO cites? -->. In any case, because status exchange positively selects one parent and negatively selects the other partner, the two effects may somewhat cancel each other out in terms of the overall distribution of resources.

@@

<!-- TODO how can I integrate Cheng and Powell? -->

@@

The experience of hybridity itself may also positively and/or negatively affect outcomes for biracial individuals. The "marginal man" theory argues that biracial individuals will find it difficult to fit in due to the dissonance of an identity that crosses strong and salient racial boundaries [@park_human_1928; @stonequist_problem_1935]. This liminal identity conflict will then result in negative mental health experiences that could feed into other negative outcomes as well. Evidence for such negative mental health experiences, however, has been mixed [@udry_health_2003; @campbell_what_2006; @cheng_multiracial_2009; @bratter_does_2011].

Marginal man theory viewed culture in monolithic terms and thus to be trapped in between two static and monolithic cultures was seen as a difficult experience. Contemporary notions of culture often use a "toolkit" approach that emphasizes the often flexible and fluid use of cultural elements <!-- TODO who to cite here? Swidler obviously, but also Dimaggio. Look at Dimaggio and Markus article -->. From this perspective, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts (i.e. code switch) <!-- TODO There are citations for this -->.

@@

Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided in some form or another. @campbell_multiracial_2009 used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races. @herman_blackwhiteother_2009 examines differences in GPA using data on high school students in California and Wisconsin in which the students reported their parents' races. Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes. @kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial Blacks and Asians<!-- With what? -->, and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites. This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parent's races.

Results in prior work tend to be imprecise and inconclusive, due to the sample size for most biracial groups. An alternative approach is to use large scale Census or American Community Survey (ACS) data to identify biracial children by the race of the biological parents in their household. This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households. Furthermore, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children and the data sources include relatively little detailed information. This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012]. <!-- TODO briefly summarize findings --> These studies provide valuable information on biracial childrens' lived experience, but do not directly measure their own outcomes.

@@

Grade retention is the practice of having students repeat a grade due to poor academic performance. The frequency of grade retention has fluctuated historically in response to changes in educational policy. Rates of grade retention increased throughout the late Twentieth century [@hauser_raceethnicity_2000] <!-- Check this because their more nuanced results do not necessarily show this --> but have declined substantially since 2004 [@warren_patterns_2014].

Prior work on grade retention has failed to show a positive effect of grade retention on later educational outcomes, and has instead found evidence of "scarring" effectsthat lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017]. The risk of grade retention also varies substantially by race and class background, with Black and Latino students at particularly high risk of grade retention relative to other groups [@warren_patterns_2014].

Because grade retention often happens very early in the K-12 system, it is one of the earliest stratification mechanisms that individuals encounter. For these same reasons, it provids an important window into understanding the placement of biracial children within a racially stratified educational environment. In the remainder of this article, I examine this placement using data from the American Community Survey.

@@

[^1]: Biracial respondents are slightly younger than monoracial respondents in the sample as a whole because interracial unions are becoming more common over time HOW STRONG COULD THE BIAS BE HERE AND WHAT CAN I DO ABOUT IT?

@@

I calculate two confidence intervals for each biracial group. Each confidence interval is in comparison to one of the constituent monoracial groups. These confidence intervals are color-coded in @fig-example to indicate the reference group. I also only draw half-intervals in the direction of the monoracial group's point estimate. Overlap in these color corresponding bars and bands indicates that the difference between the biracial group and the monoracial group is not statistically significant at $p<.05$. For example, the yellow bar shown in @fig-example for biracial Black/White students does not quite overlap with the yellow band for White students, indicating that the point estimates for these two groups are statistically distinguishable at $p<.05$. <!-- TODO provide actual p-value for comparison -->

@@

[^8]: It might seem odd to hold categorical variables at their mean, but because all categorical variables are technically entered into the models as 0/1 numeric indicator variables, their mean value can be calculated as the proportion of cases with the indicated value. This technique is frequently used in multilevel models to deal with grand and group mean centering of variables. <!-- TODO CITE raudenbush and bryk? -->

>>>>

The experiences of mixed race individuals will play an important role in the future of race in the US. A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines [@root_racially_1992]. This demographic phenomenon may have a potentially transformative effect on racial boundaries [@hochschild_creating_2012; @alba_great_2020]. While racial mixing is nothing new to US history [@morning_who_2000; @gullickson_choosing_2011], this biracial boom is notable for both its demographic size and for a post-Civil Rights environment in which historical norms of racial classification may be eroding. Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into a prevailing racial hierarchy that reinforces a system of racial inequality.

Extensive scholarship documents the identification and classification decisions made by or on behalf of multiracial individuals [@xie_racial_1997; @qian_options_2004; @roth_end_2005a; @brunsma_interracial_2005; @herman_forced_2004; @holloway_place_2009; @bratter_mother_2009; @davenport_role_2016; @liebler_boundaries_2016]. This research has generally found a surprising degree of fluidity and choice in classification decisions that is nonetheless somewhat predictable by local context, socioeconomic status and other characteristics. On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers [@kao_racial_1999; @campbell_multiracial_2009; @herman_blackwhiteother_2009]. Two important data limitations have contributed to the lack of work on this important topic.

@@

One important outcome, grade retention, is measurable and important even for young children, however. Being held back in school is a relatively common occurrence that has consequences for later educational transitions [@jimerson_metaanalysis_2001] and is disproportionately applied by race [@warren_patterns_2014]. Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience and reveals the imprint of the racial hierarchy. In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers. Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents.

@@

How might we expect the outcomes for biracial individuals to differ from their monoracial peers? In every case, the relevant comparison is how a biracial person fares relative to the two constituent groups with which their parents identify. For example, in the case of an individual with one Black and one White parent, we would compare the average outcomes for such individuals to the average outcomes for individuals with two White and two Black parents, respectively. The outcome for the Black/White group may be similar to one of the monoracial groups, it may be somewhere in between the outcomes for the two monoracial groups, or it may be higher or lower than either of the monoracial groups.

@@

Discrimination and bias by race play an important role in overall levels of racial inequality [@quillian_new_2006; @pager_sociology_2008]. Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowningly rely upon and reinforce in social interactions @quillian_new_2006]. Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers. Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into and reinforced by organizations and institutional systems [@reskin_race_2012]. In our current "colorblind" regime [@bonilla-silva_racism_2006], such systems may be formally race neutral but produce disparate impacts by race [@pager_sociology_2008].

@@

The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm concerning Black ancestry. According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries[@davis_who_1991]. @davis_who_1991 generalizes this practice into the concept of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group. By comparison a pattern of *hyperdescent* implies that individuals of mixed racial ancestry will be classified as members of the higher status group. This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry [@snipp_american_1989; @wolfe_land_2001; @gullickson_choosing_2011].

Although the hypodescent/hyperdescent paradigm is often used to theoretically frame the possible contemporary experiences of multiracial individuals, there are reasons to be cautious about the generalizability of this paradigm to both the contemporary and historical situation. Hypodescent and hyperdescent are observed in two specific historical cases involving Black ancestry and American Indian ancestry, respectively. As @iverson_regimes_2022 note, these two cases could just as parsimoniously be explained by a "dominance" model which ranks ancestries by their tendencies to be supercessive or recessive in determining identification and classification. In this framework, within the historical "racial triangle" of American colonialism, Black ancestry is supercessive, Indigenous ancestry is recessive, and White ancestry falls somewhere in between.

The development of these particular norms may also reflect the historical regime in which they were developed [@gullickson_choosing_2011; @iverson_regimes_2022]. In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us to fully understand the experiences of those of Latino and Asian ancestry due to more recent processes of racialization. Second, the historical norms for all groups may also be breaking up or transforming in the post-Civil Rights era.

@@

Racial inequality is also at least partly explained by different distributions of material, cultural, and social resources across racial groups [@conley_being_1999]. On average, socioeconomic background (e.g. parental income, wealth, and education) differs by race and such socioeconomic background is consequential for outcomes. Other less materially defined resources also vary across groups such as cultural styles, social networks, and neighborhood quality. These differences in background resources reflect the accumulation of historical processes of discrimination and, when combined with contemporary processes of discrimination and bias, magnify the overall level of racial inequality we observe.

@@

Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage [@davis_intermarriage_1941; @merton_intermarriage_1941; @fu_racial_2001]. Using education as the socioeconomic parameter, prior research has shown strong evidence of this pattern in the case of Black male/White female marriages in the US, but weaker evidence for other types of unions [@gullickson_education_2006; @fu_racial_2001; @kalmijn_educational_2010; @hou_interracial_2011].

Regardless of its overall prevalence, the overall effects of status exchange on resources are complex because it positively selects one parent and negatively selects the other partner. Therefore the two effects may somewhat cancel each other out in terms of the overall distribution of resources across parents. Complicating this issue further, research has shown that the parents of biracial children may differ from parents of monoracial children in the magnitude and types of resources that they allocate to children [@cheng_constraints_2007], even among parents of similar resources.

@@

@@

The experience of hybridity itself may also positively or negatively affect outcomes for biracial individuals. The "marginal man" theory argues that biracial individuals will find it difficult to fit in due to the dissonance of an identity that crosses strong and salient racial boundaries [@park_human_1928; @stonequist_problem_1935]. This liminal identity conflict will then result in negative mental health experiences that could feed into other negative outcomes as well. Evidence for such negative mental health experiences, however, has been mixed [@udry_health_2003; @campbell_what_2006; @cheng_multiracial_2009; @bratter_does_2011].

Marginal man theory viewed culture in monolithic and static terms and thus to be trapped in between two cultures was seen as a difficult experience. Contemporary views of culture instead emphasize the often flexible, fragmented, and fluid use of cultural bits that comprise a "toolkit" [@swidler_culture_1986; @dimaggio_culture_1997]. From this perspective, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts [@shih_multiple_2019].

@@

Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided in some form or another. @campbell_multiracial_2009 used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races. @herman_blackwhiteother_2009 examines differences in GPA using data on high school students in California and Wisconsin in which the students reported their parents' races. Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes. @kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial students with Black and Asian parentage and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites. This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parent's races.

Results in prior work tend to be imprecise and inconclusive, due to the sample size for most biracial groups. An alternative approach is to use large scale Census or American Community Survey (ACS) data to identify biracial children by the race of the biological parents in their household. This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households. Furthermore, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children and the data sources include relatively little detailed information. This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012]. These studies provide valuable information on biracial childrens' lived experience, but do not directly measure their own outcomes.

@@

Grade retention is the practice of having students repeat a grade due to poor academic performance. Rates of grade retention likely increased throughout the late Twentieth century [@frederick_have_2008] but have declined substantially since 2004 [@warren_patterns_2014].

Prior work on grade retention has failed to show a positive effect of grade retention on later educational outcomes, and has instead found evidence of "scarring" effects that lead to later negative educational outcomes, such as a higher dropout risk [@jimerson_metaanalysis_2001; @stearns_staying_2007; @andrew_scarring_2014; @hughes_effect_2017]. The risk of grade retention also varies substantially by race and class background, with Black and Latino students at particularly high risk of grade retention relative to other groups [@warren_patterns_2014].

Because grade retention often happens very early in the K-12 system, it is one of the earliest stratification mechanisms that individuals encounter. For these same reasons, it provides an important window into understanding the placement of biracial children within a racially stratified educational environment. In the remainder of this article, I examine this placement using data from the American Community Survey.

@@

[^1]: Biracial respondents are slightly younger than monoracial respondents in the sample as a whole because interracial unions are becoming more common over time which may contribute to a slight downward bias in their risk of grade retention.

@@

I calculate two confidence intervals for each biracial group. Each confidence interval is in comparison to one of the constituent monoracial groups. These confidence intervals are color-coded in @fig-example to indicate the reference group. I also only draw half-intervals in the direction of the monoracial group's point estimate. Overlap in these color corresponding bars and bands indicates that the difference between the biracial group and the monoracial group is not statistically significant at $p<.05$. For example, the yellow bar shown in @fig-example for biracial Black/White students does not quite overlap with the yellow band for White students, indicating that the point estimates for these two groups are statistically distinguishable at $p<.05$.

@@

[^8]: It might seem odd to hold categorical variables at their mean, but because all categorical variables are entered into the models as 0/1 numeric indicator variables, their mean value can be calculated as the proportion of cases with the indicated value.
